Static Analysis by Path Focusing

### 2011-05-25 report/report.tex
new implementation of Lookahead Widening.

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}
% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
% DM: et pas de depart
We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more elements in the sets\dots
\end{itemize}

@@

Abstract Interpretation proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values,
by choosing an simpler domain  $A$ called \emph{abstract domain}. 
% DM: un peu lourd et peu clair

@@

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
\widening x_2$. This guarantees the correctness of the result.
% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
% DM: une union abstraite juste avant
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i > 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i > 0
}$$

@@

$$\system{
x''_0 &=& \tilde{x} \\
x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i > 0
}$$

@@

Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
Abstract Interpretation. It is aimed at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by a convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

There have been various
work proposing a definition or a refinement of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying widening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. Most of the
cases, $n$ can be chosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\widening_T$ is defined as follows:
$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

@@

Throughout the report, a program will be represented by a control flow graph:
\begin{itemize}
\item $P$ is a finite set of control points.
\item $I_p$ is the (possibly empty)
set of initial values for each control point $p \in P$. For an initial control
point $p$, i.e a starting point of the program, $I_p$ is not empty. It is empty
in the other case.
\item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
has a semantic $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the set
of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
the transition to the set of states after the transition.
\end{itemize}

@@

\end{tikzpicture}
   \end{minipage}
   \caption{Example of program, and its associate control flow graph. This
   program comes from \cite{GopanR06}.}
\label{runningexample}
\end{figure}
\FloatBarrier

@@

Acceleration is possible when the loop is simple enough, and come to compute the
transitive closure $\tau_e^+$ of $\tau_e$, the transition function associated to
the loop.

@@

In some cases, a loop can have a non-regular behaviour, especially when it has
several paths. Some of these paths can stay impossible for a while, and become
possible after a certain number of iterations. However, the widening operator
extrapolates without taking care of these new possible paths. This can induce a
loss of precision, that Lookahead widening tries to limit.

@@

When computing the fixpoint iterations, we only consider a subset of the graph
for a while, until the convergence. In our example:
\begin{itemize}
\item Step 1:
	we only consider the path of
	the loop that is possible at the first iteration. The second path is
	ignored, and will be treated later. we compute the iterations, and we obtain
	an first invariant for this part of the graph: at point $p_1$, $x =
	y \wedge x \geq 0$.
	Then, we can do some
	narrowing steps in order to recover some precision, before adding new paths
	in the graph. At point $p_1$ after narrowing, $x=y \wedge 0 \leq x \leq 51$.
\item Step 2:
	Now, we add the new paths into our graph if they have become possible. In
	our case, the second path of the loop is now possible, because $x$ can be
	equal to $51$, so we add the path in
	the graph and compute new iterations until convergence\dots. After
	convergence, at point $p_1$, $x + y - 102 \leq 0 \wedge x
	\leq y$. We then apply a narrowing step and at $p_1$, 
$x + y - 102 \leq 0 \wedge x \leq y \wedge y \geq 0$.
\item Step 3: finally, the path from $p_4$ to $p_7$ becomes possible, so we add
it into the graph and compute the invariant for $p_7$.
\end{itemize}

@@

  \path [transition] 
		(n0) edge  [out=0, in=0]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=180, in=180]        node {} (n9);
  \path [transition] 
		(n0) edge  [out=205, in=155]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=230, in=130]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=255, in=105]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-25, in=25]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-50, in=50]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-75, in=75]            node {} (n9);
\end{tikzpicture}
\end{minipage}
\caption{One can expand the graph in the left and obtain the associate graph in
the right.}
\end{figure}
\FloatBarrier

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associate expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantic.}
\label{multigraphtransformGopan}
\end{figure}
\FloatBarrier

@@

This static single assignment form is required for the computation of the focus
path.

@@

\begin{eqnarray*}
\rho = & \displaystyle\bigwedge_{p_i \notin P_R}& \left[ (b_i = \bigvee_{(p_h,p_i) \in E}
t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i \wedge c(i,j))
\right] \wedge \\
 & \displaystyle\bigwedge_{p_i \in P_R}& \left[ (b_i^d = \bigvee_{(p_h,p_i) \in E}
 t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i^s \wedge
 c(i,j)) \right] \wedge \\
 & \displaystyle\bigwedge_{x \in \Sigma}& assign(x)
\end{eqnarray*}

@@

	\subsubsection{Expression of the reachability problem}
	\label{smtformula}
	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
	formula expressing the reachability problem defined in \ref{reachability}.

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $\rho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

\begin{itemize}
\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
\{p_0\}$.
\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
We do another SMT query: is there another path
starting in $p_0$ that makes an abstract value grow? Answer is no.
\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
is the first phase of the loop. 
The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
the same path. Then, we can do a widening/narrowing sequence. 
After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
\wedge (0 \leq x \leq 51)$.
\item Step 3: We still are in $p_1$: we ask for another focus path, and the
SMT-solver finds the path $p_1 \rightarrow p_3
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
Again, this is a self-loop, but we do not widen yet. The
image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
the convex hull of this point with the old $X_1$. A new SMT query gives the same
path, and we can now apply widening/narrowing. We obtain
$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
We can see that $X_1$ is the smallest polyhedron containing the possible values
for $(x,y)$ (see Figure \ref{result}.b).
\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
precise result of this program. After that, the SMT-solver can't find any new
path, so the analysis terminates.
\end{itemize}

@@

	\subsection{Removing Identity transitions}

@@

	$\Phi$ is the function we classically use to find an inductive invariant
	$X$,satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
	we propose to compute this invariant using another function $\Psi$, by
	removing the transitions that are actually the identity. Indeed, these
	transitions make no change to the set $X$.
	The $\Psi$ function could be expressed as follows:

@@

We consider the program depicted in figure \ref{ex9}.
Here, we would like to show that the returned value $r$ is zero.

@@

\paragraph{Path Focusing} 
The analysis of the loop starts with 
$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
widening/narrowing sequence and we find 
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT gives
another path: 
$p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
analysing this path, we obtain 
$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
program. Indeed, there is no feasible paths going through $p_6$, but
\emph{Lookahead Widening} don't know it because of the convex hull at point
$p_5$.

@@

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed after the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that is aimed at
computing these live variables after the transformation in SSA form.

@@

	We attach only one abstract value to each BasicBlock of the IR. Indeed,
	there would be a huge memory consumption if we would attach a polyhedron
	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
	and ends with the other instructions. We choose to attach our abstract value
	just after the $\Phi$-variables assignments. In this way, the
	$\Phi$-variables of the associate BasicBlock will be part of the dimensions
	of the polyhedron.

@@

	At a program point $p$, the dimensions of the associate abstract value could
	only be the numerical variables that are live at $p$. This is
	not exactly the case in our implementation, because of another 
	optimisation: we only consider
	variables that are not a linear combination of other variables.

@@

	There is a special case for $\Phi$-variables: in some cases, a
	$\Phi$-variable can be considered as a linear combination of program
	variables, when only a single incoming value is possible. This is the case
	when all but one polyhedra associated to a predecessor are at bottom.
	For instance, let say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
	predecessors. At point $p$, we have $x = \Phi(x_1, x_2,\dots,x_n)$.
	If $\forall i \in \{2,\dots,n\}, X_i = \perp$, and $X_1 \neq \perp$, then we
	can use the linear relation $x = x_1$ instead of the $\Phi$ function. In
	practical cases, this is a way to reduce the size of the abstract values.

@@

The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
represent the set of possible states of our program, we loose information when
the analysis comes across a \emph{diseq} operation.
One classical approach would be to compute a disjunctive invariant, by attaching
two different polyhedron to the BasicBlock. However, we preferred the following
technique: we transform all the operations of the form $x \neq y$ into
$x < y \vee x > y$, meaning that we separate the cases \emph{lesser than} and
\emph{greater than} into two distinct paths.

@@

In some cases, the number of iterations inside a loop during execution 
may be equals to zero. This happens if the condition of the \emph{while}
statement is false at the first time the head of loop is reached.

@@

In this example, our program behaves as follows:
\begin{itemize}
\item At point $p_1$, we first have the polyhedra $i=0$, where $n$ in
unconstrained.
\item The image of this polyhedra after the loop transition is $i \leq n \wedge
i=1$. After a convex hull, we obtain for $p_1$ $0 \leq i \leq 1$.
\item Again, we
compute the image of this polyhedron by the transition, $1 \leq i \leq 2 \wedge
i \leq n$, and after widening we
obtain at $p_1$ the polyhedron $0 \leq i$.
\item Finally, at point $p_f$, we have the polyhedron $0 \leq i \wedge i \geq
n$.
\end{itemize}

@@

	Our analyser could be extended in many ways. Indeed, it is intra-procedural,
	meaning that it does not consider that a variable may change after a
	function call. In addition, the return value of a function call is
	considered unconstrained. This limitation can be addressed by simply
	inlining the function call, provided the function is not recursive.
	There have been research to find efficient interprocedural analysis
	techniques \cite{Bourdoncle90,Bou92}.

@@

	In all experiments, the set $P_R$ is chosen equals to $P_W \cup P_{ret} \cup
	P_{init}$,
	where $P_W$ is the set of widening points. $P_{init}$ is the set of starting
	points of the program, and $P_{ret}$ is the set of points that have no
	successors.

>>>>

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}
% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
% DM: et pas de depart
We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more new elements in the
sets\dots
\end{itemize}

@@

Abstract Interpretation proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values,
by choosing a simpler domain  $A$ called \emph{abstract domain}. 
% DM: un peu lourd et peu clair

@@

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
\widening x_2$. This guarantees the correctness of the result.
% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
% DM: une union abstraite juste avant
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i \geq 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i \geq 0
}$$

@@

$$\system{
x''_0 &=& \tilde{x} \\
x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i \geq 0
}$$

@@

Linear Relation Analysis \cite{CH78} (also denoted LRA) is a direct application
of
Abstract Interpretation. It is aimed at computing an upper approximation of the
reachable states of a program containing numerical variables. The set of
possible assignments for the numerical variables is abstracted by a convex
polyhedron. This technique discovers invariant linear relations between
the numerical variables at each control point of a program.

@@

There have been various
work proposing a definition or a refinement of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying widening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. In most of
the cases, $n$ can be chosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\widening_T$ is defined as follows:
$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

@@

Throughout the report, a program will be represented by a control flow graph:
\begin{itemize}
\item $P$ is a finite set of control points.
\item $I_p$ is the (possibly empty)
set of initial values for each control point $p \in P$. For an initial control
point $p$, i.e a starting point of the program, $I_p$ is not empty. It is empty
in the other case.
\item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
has a semantics $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the
set
of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
the transition to the set of states after the transition.
\end{itemize}

@@

\end{tikzpicture}
   \end{minipage}
   \caption{Example of program, and its associated control flow graph. This
   program comes from \cite{GopanR06}.}
\label{runningexample}
\end{figure}
\FloatBarrier

@@

Acceleration is possible when the loop is simple enough, and comes to compute
the transitive closure $\tau_e^+$ of $\tau_e$, the transition function
associated to the loop.

@@

In some cases, a loop can have a non-regular behaviour, especially when it has
several paths. Some of these paths can stay impossible for a while, and become
possible after a certain number of iterations. However, the widening operator
extrapolates without taking care of these new possible paths. This can induce a
loss of precision, that \emph{Lookahead widening} tries to limit.

@@

When computing the fixpoint iterations, we only consider a subset of the graph
for a while, until the convergence. In our example:
\begin{itemize}
\item Step 1:
	we only consider the path of
	the loop that is possible at the first iteration. The second path is
	ignored, and will be treated later. We compute the iterations, and we obtain
	an first invariant for this part of the graph: at point $p_1$, $x =
	y \wedge x \geq 0$.
	Then, we can do some
	narrowing steps in order to recover some precision, before adding new paths
	in the graph. At point $p_1$ after narrowing, $x=y \wedge 0 \leq x \leq 51$.
\item Step 2:
	Now, we add the new paths into our graph if they have become possible. In
	our case, the second path of the loop is now possible, because $x$ can be
	equal to $51$, so we add the path in
	the graph and compute new iterations until convergence\dots. After
	convergence, at point $p_1$, $x + y - 102 \leq 0 \wedge x
	\leq y$. We then apply a narrowing step and at $p_1$, 
$x + y - 102 \leq 0 \wedge x \leq y \wedge y \geq 0$.
\item Step 3: finally, the path from $p_4$ to $p_7$ becomes possible, so we add
it into the graph and compute the invariant for $p_7$.
\end{itemize}

@@

  \path [transition] 
		(n0) edge  [out=0, in=0]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=180, in=180]        node {} (n9);
  \path [transition] 
		(n0) edge  [out=205, in=155]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=230, in=130]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=255, in=105]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-25, in=25]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-50, in=50]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-75, in=75]            node {} (n9);
\end{tikzpicture}
\end{minipage}
\caption{One can expand the graph in the left and obtain the associated graph in
the right.}
\end{figure}
\FloatBarrier

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associated expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantic.}
\label{multigraphtransformGopan}
\end{figure}
\FloatBarrier

@@

This static single assignment form is used for the computation of the focus
path.

@@

\begin{eqnarray*}
\rho = & \displaystyle\bigwedge_{p_i \notin P_R}& \left[ \left(b_i = \bigvee_{(p_h,p_i) \in E}
t_{h,i}\right) \wedge \left( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i \wedge
c(i,j)\right)
\right] \wedge \\
 & \displaystyle\bigwedge_{p_i \in P_R}& \left[ \left(b_i^d = \bigvee_{(p_h,p_i) \in E}
 t_{h,i}\right) \wedge \left( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i^s \wedge
 c(i,j)\right) \right] \wedge \\
 & \displaystyle\bigwedge_{x \in \Sigma}& assign(x)
\end{eqnarray*}

@@

	\subsubsection{Expression of the reachability problem}
	\label{smtformula}
	The SMT-formula $\rho$ express the semantics of the program. We still need
	to create the
	formula expressing the reachability problem defined in \ref{reachability}.

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $\rho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associated abstract values $X_0$, $X_1$ and $X_7$.

\begin{itemize}
\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
\{p_0\}$.
\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
We do another SMT query: is there another path
starting in $p_0$ that makes an abstract value grow? The answer is no.
\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
is the first phase of the loop. 
The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
the same path. Then, we can do a widening/narrowing sequence. 
After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
\wedge (0 \leq x \leq 51)$.
\item Step 3: We still are in $p_1$: we ask for another focus path, and the
SMT-solver finds the path $p_1 \rightarrow p_3
\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
Again, this is a self-loop, but we do not widen yet. The
image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
the convex hull of this point with the old $X_1$. A new SMT query gives the same
path, and we can now apply widening/narrowing. We obtain
$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
We can see that $X_1$ is the smallest polyhedron containing the possible values
for $(x,y)$ (see Figure \ref{result}.b).
\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
precise result of this program. After that, the SMT-solver can't find any new
path, so the analysis terminates.
\end{itemize}

@@

	\subsection{Removing identity transitions}

@@

	$\Phi$ is the function we classically use to find an inductive invariant
	$X$, satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
	we propose to compute this invariant using another function $\Psi$, by
	removing the transitions that are actually the identity. Indeed, these
	transitions make no change to the set $X$.
	The $\Psi$ function could be expressed as follows:

@@

We consider the program depicted in Figure \ref{ex9}, page \pageref{ex9}.
Here, we would like to show that the returned value $r$ is zero.

@@

\paragraph{Path Focusing} 
The analysis of the loop starts with 
$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
widening/narrowing sequence and we find 
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT-solver
gives another path: 
$p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
analysing this path, we obtain 
$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
program. Indeed, there is no feasible paths going through $p_6$, but
\emph{Lookahead Widening} don't know it because of the convex hull at point
$p_5$.

@@

Another example of information that was missing is the set of live variables.
Indeed, in LLVM, live variables computation is processed during the register
allocation. Then, the data structure we work on doesn't have live variables
informations yet. To solve this problem, we coded our own pass, that is aimed at
computing these live variables after the transformation in SSA form.

@@

	We attach only one abstract value to each BasicBlock of the IR. Indeed,
	there would be a huge memory consumption if we would attach a polyhedron
	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
	and ends with the other instructions. We choose to attach our abstract value
	just after the $\Phi$-variables assignments. In this way, the
	$\Phi$-variables of the associated BasicBlock will be part of the dimensions
	of the polyhedron.

@@

	At a program point $p$, the dimensions of the associated abstract value could
	only be the numerical variables that are live at $p$. This is
	not exactly the case in our implementation, because of another 
	optimisation: we only consider
	variables that are not a linear combination of other variables.

@@

	There is a special case for $\Phi$-variables: in some cases, a
	$\Phi$-variable can be considered as a linear combination of program
	variables, when only a single incoming value is possible. This is the case
	when all but one polyhedra associated to a predecessor are at bottom.
	For instance, let's say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
	predecessors. At point $p$, we have $x = \Phi(x_1, x_2,\dots,x_n)$.
	If $\forall i \in \{2,\dots,n\}, X_i = \perp$, and $X_1 \neq \perp$, then we
	can use the linear relation $x = x_1$ instead of the $\Phi$ function. In
	practical cases, this is a way to reduce the size of the abstract values.

@@

The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
represent the set of possible states of our program, we loose information when
the analysis comes across a \emph{diseq} operation.
One classical approach would be to compute a disjunctive invariant, by attaching
two different polyhedron to the BasicBlock. However, we preferred the following
technique: we transform all the operations of the form $x \neq y$ into
$x < y \vee x > y$, meaning that we separate the cases \emph{less than} and
\emph{greater than} into two distinct paths.

@@

In some cases, the number of iterations inside a loop during execution 
may be equal to zero. This happens if the condition of the \emph{while}
statement is false at the first time the head of loop is reached.

@@

In this example, our program behaves as follows:
\begin{itemize}
\item At point $p_1$, we first have the polyhedron $i=0$, where $n$ in
unconstrained.
\item The image of this polyhedron after the loop transition is $i \leq n \wedge
i=1$. After a convex hull, we obtain for $p_1$ $0 \leq i \leq 1$.
\item Again, we
compute the image of this polyhedron by the transition, $1 \leq i \leq 2 \wedge
i \leq n$, and after widening we
obtain at $p_1$ the polyhedron $0 \leq i$.
\item Finally, at point $p_f$, we have the polyhedron $0 \leq i \wedge i \geq
n$.
\end{itemize}

@@

	Our analyser could be extended in many ways. Indeed, it is intra-procedural,
	meaning that it does not consider that a variable may change after a
	function call. In addition, the return value of a function call is
	considered unconstrained. This limitation can be addressed by simply
	inlining the function call, provided the function is not recursive.
	There have been research to find efficient interprocedural analysis
	techniques \cite{Bourdoncle90,Bou92,jeannet09b}.

@@

	In all experiments, the set $P_R$ is chosen equal to $P_W \cup P_{ret} \cup
	P_{init}$,
	where $P_W$ is the set of widening points. $P_{init}$ is the set of starting
	points of the program, and $P_{ret}$ is the set of points that have no
	successors.
